---
title: "Numeric , Arithmetic , Assignment and Vectors"
date: "25-12-2024"
output: pdf_document
---

In this we introduces the use of R for basic mathematics, storing results for later use, and vectors, which are essential in R. In this we emphasizes that much of R functionality is designed with vector operations in mind.

## Arithmetic functions
```{r}
10+10   # addition  
10/10   # division
10-10   # subtraction
10*10   # multiplication
sqrt(10)# square root 
10^2    # power
```

## logarithm functions
```{r}
log(10)   # log
exp(10)   # exponent
```
## E-notation
```{r}
1e10
1e10/100  # ten raise to the power
```
## Assignment and Variables

The assignment operator (<-) is used to store values in variables.This stores
the value on the right side of the operator to the object on the left.

```{r}
# objectifying the values
x<-10/3   # assign a value to a x
x
```
```{r}
# Once a value is stored in a variable, it can be used in other calculations.
x+3
x*8
x/3
```
## Vectors

Vectors are one-dimensional arrays that can hold multiple values of the same 
data type.The c() function (which stands for `concatenate`) is used to create
vectors.

```{r}
my_vec<-c(1,2,3,4,5)
my_vec
```
## vector arithmetic

Arithmetic operations can be performed on vectors.When an arithmetic operation
is performed on two vectors, R performs the operation element-by-element.

```{r}
v1<-c(2,4,6,8,5)
v2<-c(3,5,7,9,11)

v1-v2
v1/v2
v1*v2
v1+v2
```
## Recycling

If two vectors are of different lengths, R will recycle the shorter vector to
match the length of the longer vector.

```{r}
v1
v1+1
v1 + c(1,2)  # c(1,2) was recycled to c(1,2,1,2) to match the length of v1. 
```
## Regular Sequences

R provides functions for generating regular sequences of numbers.The colon operator (:) is used to create a sequence of integers.

```{r}
y<-1:20  # generate range of no. form 1 to 20.
y
```
```{r}
# The seq() function creates more complex sequences.
seq(from=2,to=10)               # no. seq range from 2 to 10
seq(from=2,to=10,by=2)          # no. seq range by gap of 2.
seq(from=2,to=10,length.out=9)  # no. seq range of length 9.
```

## Sub-setting Vectors

The square brackets ([ ]) are used to extract elements from a vector.The index
or indices of the element(s) to be extracted are placed inside the brackets.

```{r}
# value extraction from vectors.
v1
v1[2]
v1[c(2,5)]
```
```{r}
# To omit elements, use negative index numbers inside the square brackets.
v1[-2]
v1[-c(2,4)]
```
```{r}
# You can use logical vectors (contain only TRUE or FALSE) to extract elements from vectors.
v1
v1[c(TRUE,FALSE,TRUE,TRUE,FALSE)] # false means explode it form vector.
```

## Vector Functions

R provides many functions for working with vectors.Some Useful Functions are as follows:

```{r}
v2
length(v2)
sum(v2)
prod(v2)
sqrt(v2)
sort(v2)
```

Vectorized Functions meaning that they can operate on entire vectors at once.

```{r}
v3<-c(2.3456,3.5678,3.456,1.267)
round(v3,2) # round to two decimal places

log(v3)     # log of a vector that is v3.
```






