---
title: "Non Numeric Values"
output: pdf_document
---
In this we are generally focusing on working with non-numeric data types, an essential skill for effective statistical programming in R. These data types play a crucial role when dealing with real-world data sets and become especially important when working with more complex R programming.The chapter covers three important non-numeric data types:

1. Logical: representing TRUE/FALSE values.
2. Characters: representing textual data (strings).
3. Factors: representing categorical data.

## Logical Values
Logical values, often referred to as logicals, are fundamental to programming. They are based on the principle that a logical-valued object can be either TRUE or FALSE.  These values can represent concepts like yes/no, one/zero, or satisfied/not satisfied, and they have numerous applications in programming.

Reserved Values: In R, the terms TRUE and FALSE are reserved specifically for representing logical values. This means you cannot use these terms to name objects or assign them different values.

Unreserved Equivalents:R also allows you to use T and F as shorthand for TRUE and FALSE.However, T and F are not reserved words in R. This means a user could potentially redefine them, leading to unexpected behavior. For this reason, it's generally recommended to use TRUE and FALSE for clarity and to avoid potential errors.
```{r}
x<-TRUE
x
```
```{r}
y<-F
y
```
```{r}
Logical <- c(T,F,F,F,T,F,T,T,T,F,T,F)
Logical

length(Logical)
```
```{r}
Logical_1<- matrix(data=Logical,nrow=3,ncol=4)
Logical_1
```
## Relational Operators
Relational operators are used to compare values,returning a logical result (TRUE or FALSE) based on the comparison.That are ==(Equal),=(Not Equal to),>(Greater than),<(Less than),>=(Greater than Equal to),<=(Less than Equal to),any() checks if at least one element in a logical vector is TRUE and all() checks if all elements in a logical vector are TRUE.
```{r}
1==2  # Equal
1!=2  # Not Equal
1>=2  # Greater than Equal to
1<=2  # Less than Equal to
```
```{r}
x_1 <- c(3,2,1,4,1,2,1,-1,0,3)
x_2 <- c(4,1,2,1,1,0,0,3,0,4)

x_1==x_2    # Equal
x_1!=x_2    # Not Equal
x_1>=x_2    # Greater than Equal to
x_1<=x_2    # Less than Equal to
```
# Logical Operators
Logical operators are used to combine or modify logical values.The key operators are:

1. **AND** (&, &&):Returns TRUE if both operands are TRUE.& is vectorized,meaning it operates on entire vectors element-wise.&& is element-wiseand only compares the first elements of vectors. It is primarily used in conditional statements (if, while).
```{r}
TRUE & TRUE   
TRUE && FALSE
```
2. **OR** (|, ||): Returns TRUE if at least one operand is TRUE.| is vectorized.|| is element-wise and mainly used in conditional statements.
```{r}
TRUE | FALSE 
FALSE || TRUE
```
3. **NOT** (!):  Negates a logical value, flipping TRUE to FALSE and vice versa.
```{r}
!TRUE  
!FALSE
```
# Logical as Numeric   
R can interpret logical values as numeric values:TRUE is equivalent to 1 FALSE is equivalent to 0.This allows you to use logical values in arithmetic calcula- tions and with functions that expect numeric input.
```{r}
TRUE + TRUE + FALSE   
sum(c(TRUE, FALSE, TRUE))
```
# Logical Sub-setting and Extraction
A key use of logical values is for sub-setting and extracting elements from vectors and other data structures.You can use a logical vector as an index inside square brackets ([ ]) to select elements. Only elements corresponding to TRUE values in the logical vector will be extracted.
```{r}
myvec <- c(5,-2.3,4,4,4,6,8,10,40221,-8) # for getting specific values.
myvec[c(F,T,F,F,F,F,F,F,F,T)]

myvec[myvec<0]
```
The logical vector used for sub-setting must be the same length as the vector being subsetted, although recycling can occur if the logical vector is shorter.
```{r}
A <- matrix(c(0.3,4.5,55.3,91,0.1,105.5,-4.2,8.2,27.9),nrow=3,ncol=3)
A
A[c(T,F,F),c(F,T,T)]
A<1
A[A<1] <- -7
```